---
title: "Data processing"
author: "Health Data Science Unit"
date: "`r date()`"
output: html_document
---


```{r include=FALSE}
library(Seurat)
library(tidyverse)
source('../scripts/pipeline.R')
#library(harmony)

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE, 
                      cache = TRUE,
                      fig.width = 7,
                      fig.height = 7, 
                      fig.align = TRUE)
```

## Himmer data


## Loading data

```{r loading_data}
himmer <- read.csv('../data/nina_thimme_raw_counts.csv', 
                header = TRUE)
r_names <- himmer$X
himmer <- select(himmer, -X)
himmer_mtx <- apply(himmer, 2, as.numeric)
rownames(himmer_mtx) <- gsub('__chr.*', '', r_names) 

himmer_seu <- CreateSeuratObject(counts = himmer_mtx, project = 'himmer', min.cells = 1, assay = 'rna')
himmer_seu <- st_workflow(himmer_seu, n_features = 1000, n_pca_dims = 100)
```


## Data vis

```{r}
DimPlot(himmer_seu, group.by = 'orig.ident', reduction = 'umap') 
```

## Batch correction

### Splitting subsets by patient

```{r }
patients <- himmer_seu$orig.ident
himmer_df <- as.data.frame(t(himmer_mtx))
himmer_sp <- split(himmer_df, f = patients) 
himmer_sp <- lapply(himmer_sp, function(x) t(x) )
#saveRDS(himmer_sp, '../data/himmer_splitted.rds')
```


## Loading Miller data

```{r}
barcodes_url <- 'https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE122712&format=file&file=GSE122712%5Fbarcodes%2Etsv%2Egz'
genes_url <- 'https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE122712&format=file&file=GSE122712%5Fgenes%2Etsv%2Egz'
counts_url <- 'https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE122712&format=file&file=GSE122712%5Fmatrix%2Emtx%2Egz' 
if (! file.exists('../data/miller2019/barcodes.tsv')){
        download.file(barcodes_url, destfile = '../data/miller2019/GSE122712_barcodes.tsv.gz')
        gunzip('../data/miller2019/GSE122712_barcodes.tsv.gz',
               destname = '../data/miller2019/barcodes.tsv', 
               remove = FALSE)
}
if (! file.exists('../data/miller2019/genes.tsv')) {
        download.file(genes_url, destfile = '../data/miller2019/GSE122712_genes.tsv.gz')
        gunzip('../data/miller2019/GSE122712_genes.tsv.gz', 
               destname = '../data/miller2019/genes.tsv', 
               remove = FALSE)
}
if (! file.exists('../data/miller2019/matrix.mtx.gz')) {
        download.file(counts_url, destfile = '../data/miller2019/matrix.mtx.gz')
        gunzip('../data/miller2019/matrix.mtx.gz', 
               destname = '../data/miller2019/matrix.mtx', 
               remove = FALSE)
}
```

## Standardizing gene names

* Gene names should be in upper case in order to align with human data.

```{r seurat_object, eval=FALSE}
## Run once in order to create ../data/miller2019_upper_Case/

#dir.create('../data/miller2019_upper_Case')
#file.copy('../data/miller2019/matrix.mtx', '../data/miller2019_upper_Case/')
#file.copy('../data/miller2019/barcodes.tsv', '../data/miller2019_upper_Case/')
genes_lc <- read.table('../data/miller2019/genes.tsv', header = FALSE)
genes_lc$V2 <- toupper(genes_lc$V2)
write_tsv(genes_lc, '../data/miller2019_upper_Case/genes.tsv', col_names = FALSE)
```

```{r}
miller_mtx <- Read10X('../data/miller2019_upper_Case/')
miller <- CreateSeuratObject(counts = miller_mtx, project = 'lcmv', assay = 'rna', min.cells = 1, min.features = 200)
```

## Clustering Miller data

### Seurat standard preprocessing

```{r miller_seurat_std_preprocessing}
miller <- NormalizeData(miller)
miller <- FindVariableFeatures(miller, selection.method = 'vst', nfeatures = 3000)
miller <- ScaleData(miller, verbose = FALSE)
miller <- RunPCA(miller, npcs = 30, verbose = FALSE)
miller <- RunUMAP(miller, reduction = "pca", dims = 1:30)
miller <- FindNeighbors(miller, dims = 1:30, verbose = FALSE)
miller <- FindClusters(miller, resolution= 0.1, verbose = FALSE)
```

### Cluster annotation

```{r annotate_clusters}
miller$orig.ident <- miller$seurat_clusters
miller_ann <- data.frame('cluster' = miller$orig.ident,
                      'cell_type' = plyr::mapvalues(x = miller$orig.ident,
                                                    from = as.factor(c(3,0,4,2,1)), 
                                                    to = c('Proliferating', 'Effector', 'Naive',
                                                           'Progenitor Ex', 'Terminally Ex')
                                                    )
)
miller <- AddMetaData(miller, metadata = miller_ann)
saveRDS(miller_ann, '../data/miller_annotations.rds')
miller.vis <- DimPlot(miller, group.by = 'cell_type', reduction = 'umap', label = TRUE) + NoLegend() + ggtitle('RNA')
miller.vis
```



## Merging Miller and Himmer datasets

* In addition to using upper case in the Miller data, gene names must be processed in Himmer databases to remove '_chromosome#' tag.

```{r merging_Himmer_Miller}
## checking no empty intersection of genes
#length( intersect( rownames(himmer_sp$DW), rownames(miller_mtx)))

## merging miller and himmer in a list
merged_himmer_miller <- c('Miller_dataset'=miller_mtx, himmer_sp)
saveRDS(merged_himmer_miller, '../data/merged_himmer_miller.rds')
```

